---
title: "Hw1"
output: html_document
date: "2025-09-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2. Vectors & Factors
## 2.1 Named numeric vector, attributes

```{r}
wts <- c(10, 20, 30, 40, 50)
names(wts) <- c("a", "b", "c", "d", "e")

attr(wts, "units") <- "kg"

typeof(wts)
length(wts)
names(wts)
attributes(wts)
```

## 2.2 Special values & coercion

```{r}
v <- c(1, NA, NaN, Inf, "2")
typeof(v)
```

By default, R will coerce mixed data types of numbers and characters into only being characters. That's why `typeof(v)` returns the value "character".

```{r}
v_num <- as.numeric(v)
finite_vals <- v_num[is.finite(v_num)]
median_val <- median(finite_vals)
print(median_val)

v_clean <- v_num
v_clean[!is.finite(v_clean)] <- median_val
v_clean[is.na(v_clean)] <- median_val
v_clean
```
## 2.3 Factors

In the mtcars dataset, 0 is automatic and 1 is manual. In the displayed table, there are 19 automatics and 13 manuals.

```{r}
table(mtcars$am)
am_f <- factor(mtcars$am, 
               levels = c(0, 1), 
               labels = c("Automatic", "Manual"))
am_f
```
```{r}
am_f <- relevel(am_f, ref = "Manual")
levels(am_f)
table(am_f)

```
# 3 Data Frames & Coercion


## 3.1 Data frame construction

Create a dataframe called df_small with columns: id (1:6), group (a two-level factor variable A/B), val (a numeric variable with one NA value). Ensure classes are respected.
```{r}
id <- 1:6
group <- factor(c("A", "B", "A", "B", "A", "B"))
val <- c(8, 1, NA, 15, 25, 13)
df_small <- data.frame(id, group, val)
```

## 3.2 Coercion rules (vector vs data frame)

Bind `df_small$group` and `df_small$val` into a single vector using c(...). Explain the resulting type and why. Then show a structure that preserves numeric values and group labels for later grouping.
```{r, results='hide'}
binded_vector <- c(df_small$group,df_small$val)
typeof(binded_vector) #Returns "double"
binded_df <- data.frame(df_small$group,df_small$val) #Preserves values and labels
```


# 4 Subsetting (12 pts)

```{r}
```

## 4.1 By position, name, logical

```{r}
```

## 4.2 Negative indices via matching

```{r}
```

## 4.3 List subsetting differences

```{r}
```


# 5 Matrices & Arrays (18 pts)

```{r}
```

## 5.1 Model matrix with intercept

```{r}
```

## 5.2 Least Squares via matrix operations

```{r}
```

## 5.3 AIC

```{r}
```

# 6 Model Selection

```{r}
```

## 6.1 Best model per dimension

```{r}
```
